Neutral ISM, Lyman-Alpha and Lyman-continuum in nearby starburst Haro 11 \footnote{Based on observations with HST-COS, program GO 13017, PI Heckman}

### 2017-01-26 ToSubmit/RiveraThorsen.tex
Second revision submitted, hopefully final revision

\begin{abstract} \textbf{ Star forming galaxies are believed to be a major
		source of Lyman Continuum (LyC) radiation responsible for
		reionizing the early Universe. Direct observations of escaping
		ionizing radiation have however been few and with low escape
		fractions. In the local Universe, only ~10 emitters have been
		observed, with typical escape fractions of a few percent. The
		mechanisms regulating this escape need to be strongly evolving
		with redshift in order to account for the Epoch of Reionization.
		Gas content and star formation feedback are among the main
		suspects, known to both regulate neutral gas coverage and evolve
		with cosmic time. In this paper, we reanalyze HST-COS
		spectrocopy of the first detected local LyC leaker, Haro 11. We
		examine the connection between LyC leakage and Lyman-$\alpha$
		line shape, and feedback-influenced neutral ISM properties like
		kinematics and gas distribution. We discuss the two extremes of
		an optically thin, density bounded ISM and a riddled, optically
		thick, ionization bounded ISM, and how Haro 11 fits into their
		theoretical predictions.  We find that the most likely ISM model
		is a clumpy neutral medium embedded in a highly ionized medium
		with a combined covering fraction of unity and a residual
		neutral gas column density in the ionized medium high enough to
		be optically thick to Lyman-$\alpha$, but low enough to be at
		least partly transparent to Lyman continuum and undetected in Si
		II. This suggests that SF feedback and galaxy-scale interaction
		events play a major role in opening passageways for ionizing
radiation through the neutral medium.} \end{abstract}

%%% SUBMITTED ORIGINAL VERSION
% Star forming galaxies are believed to be a major source of ionizing
% radiation responsible for reionizing the early Universe. Direct
% observations of escaping ionizing radiation have however been few and
% with escape fractions far below what would be necessary to account for
% the ionization history of the Universe. In the local Universe, only a
% few handfuls of emitters have been observed with typical escape
% fractions of a few percent. It seems the mechanisms regulating this
% escape need to be strongly evolving with redshift, if galaxies are to be
% held responsible for reionization. Gas mass, star formation feedback and
% thus star formation activity are among the main suspect parameters,
% known to both regulate neutral gas coverage and evolve with cosmic time.
% In this paper, we present a reanalysis of far-UV HST-COS spectra of the
% first detected Lyman continuum leaker in the local Universe, Haro 11, in
% which we examine the connections between Lyman continuum leakage and
% Lyman-$\alpha$ line shape and feedback-influenced neutral ISM properties
% like kinematics, geometry and column density. We discuss the two
% extremes of an optically thin, density bounded ISM and a riddled,
% optically thick, ionization bounded ISM and how Haro 11 fits into the
% theoretical predictions made from these models. We find that the most
% likely ISM model for Haro 11 knot C is one of clumpy neutral medium
% embedded in a highly ionized inter-clump medium with a combined covering
% fraction of unity and a residual neutral gas column density in the
% ionized medium of
% $5 \times 10^{13} \leq N_{\rm HI} \leq 4 \times 10^{15}$ cm⁻²; high
% enough to be optically thick to Lyman-$\alpha$, but low enough to be at
% least partly transparent to Lyman continuum and undetected in
% \ion{Si}{2}.

@@

\begin{figure}
\centering
\includegraphics[width=3.500in]{./HISLISProfiles.pdf}
\caption{The \ion{Si}{2} and other LIS (\textbf{upper}) and \ion{Si}{4}
(\textbf{lower}) profiles included in this study. Dashes denote data
points masked out due to foreground contamination. \textbf{Data in this figure has
been smoothed to better show the line shapes, but all computations have been
carried out on the unsmoothed data.} Note the
high-velocity component in both \ion{Si}{4} lines, extending out to
$\sim 850$~km~s$^{-1}$.}\label{fig:SingleLines}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3.500in]{./LyACoverfracs.pdf}
\caption{\textbf{Upper panel}: Ly$\alpha$ profile of Haro 11 C, in
approximate units of the surrounding continuum level. \textbf{The full line
indicates the measured values smoothed by a 5 px. flat kernel}; surrounding shading
encloses the $\pm 1 \sigma$ error band (mostly comparable in width to
the line width). \textbf{Middle panel}: Black steps show the averaged
LIS line profile, smoothed by a 5px kernel. Surrounding gray shading
denotes the $\pm 1 \sigma$ confidence band. \textbf{Lower panel}: Same
as middle panel, but for the \ion{Si}{4} transitions.
}\label{fig:HisLisLya}
\end{figure}

The picture is what we would expect from a LyC leaker: \textbf{The
neutral absorption features are weak, and the ionized features are
strong, revealing a   highly ionized medium
in front of the central cluster} . This is also in good agreement with the
findings of \citet{Pardy2016arXiv}, who conclude that the galaxy has
about twice as much ionized as neutral gas. In addition to this, it is
interesting to note the close similarity in shape between the
\ion{Si}{2} and \ion{Si}{4} line profiles, indicating that they likely
represent two different phases in the same higher density regions. These
regions will likely be ionized on the side facing the central cluster,
being photoionized by its hot, massive stars. This also implies that the
nebular emission outflow found in Rivera-Thorsen et al, submitted,
at least partially traces the same gas as the \ion{Si}{4} absorption
here and, by extension, also the neutral medium. Looking at
fig.~\ref{fig:SingleLines}, we see that maximum absorption and thus the
largest amount of gas in a single velocity bin for both phases is found
at around $v \approx -50$ km s⁻¹\textbf{. Interestingly, a component was also
found in nebular emission at this velocity in the observations of 
Rivera-Thorsen et al., submitted}. The outflow velocity
$v_{\text{int}}$ is at 149 ± 7 km s⁻¹, fully consistent with the
findings of \citet{Heckman2015} and \citet{Alexandroff2015}.
Interestingly, \citet{Sandberg2013} found from the neutral Sodium
resonance absorption doublet $\lambda \lambda 5889.95,5895.92$ (NaD) an
overall, weak \emph{redshift} of $v = 32$ km s⁻¹. From
e.g.~fig.~\ref{fig:HisLisLya}, it is evident that while neutral gas
\emph{is} present at these velocities, the velocity is at odds with the
integrated velocity of $v = -149$ km s⁻¹ found in this work. We note
that \ion{Na}{1} has a very low ionization potential of only $\sim 5$
eV, meaning that these atoms may well only be present in the densest
and/or dustiest regions, in which Sodium is shielded from ionization.
This is in agreement with the finding of \citet{Sandberg2013} that NaD,
despite being a strong transition, shows absorption of only $\sim 95\%$
of continuum level and is mostly found in small, optically thick clouds.
This suggests that the NaD is tracing only the densest and/or dustiest
regions which are more slowly accelerated by star formation feedback
than the surrounding, more dilute medium. In fig.~\ref{fig:apert}, dusty
regions in the aperture are apparent E and W of knot C. \textbf{Seen in the
better resolution of fig. 1 in \citet{Adamo2010}, these regions seem
like they might be connected by a narrow dust lane partially covering
the background source. We speculate that the NaD absorption of
\citet{Sandberg2013} might be associated with this.}

Also the absorption feature in the Lyman $\alpha$ profile in the upper
panel seems to morphologically follow the shape of the metal lines,
indicating that radiative transfer effects are modest, indicative of a
fairly low column density of \ion{H}{1} around line center. We find a
Lyman $\alpha$ peak velocity of
$v_{\rm peak}^{\mathrm{Ly}\alpha} = 158 \pm 1$ km s⁻¹ \textbf{relative to} the 
systemic velocity found by \citet{Sandberg2013}, derived from nebular emission
lines in the region around knot C. This velocity, as it is also
discussed in \citet{Verhamme2015}, is just consistent with their
theoretical predictions for a density-bounded, low-column density
system, albeit on the upper limit of their allowed range.

@@

However, the \textbf{strong} riddled ionization bounded scenario 
\textbf{-- i.e. consisting of only a fully neutral and a fully ionized phase --} 
is easily
ruled out since the Ly$\alpha$ profile does not have any appreciable
emission component at zero velocity. We therefore expect a residual
neutral fraction to remain in the ionized phase; a fraction which has a
column density high enough to block Ly$\alpha$ efficiently at line
center, but low enough to be at least part transparent to LyC
radiation and undetected in \ion{Si}{2}.

@@

The existence of a diffuse neutral component being present in the
ionized medium seems consistent with what is found in \textbf{Lyman Break
Analogs and Green Pea galaxies \citep{Heckman2011, Henry2015}}. 
These galaxies have sometimes very low
LIS absorption depths, which is usually indicative of low covering
fractions and high porosity of the neutral medium - and yet they find
covering fractions of near unity of \ion{H}{1} from Lyman $\beta$
absorption, indicating column densities of $N \gtrsim 10^{16}$ cm$^{-2}$. Like in
the case of Haro 11, this indicates that a non-negligible neutral
component must be present in the ionized phase, although it may very
well be of such low densities and metallicity that metal absorption from
this medium is undetectable.

@@

In this work, we have re-analyzed an archival FUV HST-COS spectrum to
investigate the kinematics and geometry of both the hot, ionized and
cold, neutral ISM along the \textbf{line of sight to a strongly star forming
knot suspected of being the source of the observed LyC leakege from
the galaxy}. We have used the Apparent Optical Depth method to compute column 
densities and velocity-binned covering fractions of the gas and compared these 
results to the extreme cases of either an optically thin, density bounded
neutral medium, or a riddled, optically thick, ionization bounded
neutral medium. Assuming that the LyC emission previously
observed from this galaxy indeed does originate from knot C, we find
that the observations are not compatible with the latter case, since the
characteristic, bright Lyman-$\alpha$ emission spike at line center is
absent in this spectrum. Furthermore, the observations are not
consistent with an optically thin, density bounded neutral medium.

@@

\textbf{It is interesting to note that Haro 11 is relatively on neutral 
gas poor, and has a very strong ongoing starburst episode. If we were to see a 
case in which neutral gas depletion and/or ionization provided a 
density-bounded scenario, this would be a likely place to find it. 
Still, it seems that dynamical effects of feedback and galaxy-scale 
interaction plays the greater part in peeling away the neutral gas from the 
central starbursts and opening passages for Ly$\alpha$ and LyC to 
escape. Given that star formation and merger activity is significantly stronger 
at $z\gtrsim 1$ than in the local Universe, and that Dark Matter potential wells
were shallower, these mechanisms could be even more important in that era. 
one could speculate that this could provide a piece of the puzzle of which
sources have driven the epoch of reionization.}

>>>>

\begin{abstract}

Star forming galaxies are believed to be a major source of Lyman Continuum (LyC)
radiation responsible for reionizing the early Universe. Direct observations of
escaping ionizing radiation have however been few and with low escape fractions.
In the local Universe, only ~10 emitters have been observed, with typical escape
fractions of a few percent. The mechanisms regulating this escape need to be
strongly evolving with redshift in order to account for the Epoch of
Reionization. Gas content and star formation feedback are among the main
suspects, known to both regulate neutral gas coverage and evolve with cosmic
time. In this paper, we reanalyze HST-COS spectrocopy of the first detected
local LyC leaker, Haro 11. We examine the connection between LyC leakage and
Lyman-$\alpha$ line shape, and feedback-influenced neutral ISM properties like
kinematics and gas distribution. We discuss the two extremes of an optically
thin, density bounded ISM and a riddled, optically thick, ionization bounded
ISM, and how Haro 11 fits into their theoretical predictions.  We find that the
most likely ISM model is a clumpy neutral medium embedded in a highly ionized
medium with a combined covering fraction of unity and a residual neutral gas
column density in the ionized medium high enough to be optically thick to
Lyman-$\alpha$, but low enough to be at least partly transparent to Lyman
continuum and undetected in Si II. This suggests that SF feedback and
galaxy-scale interaction events play a major role in opening passageways for
ionizing radiation through the neutral medium.

\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=3.500in]{./HISLISProfiles.pdf}
\caption{The \ion{Si}{2} and other LIS (\textbf{upper}) and \ion{Si}{4}
(\textbf{lower}) profiles included in this study. Dashes denote data
points masked out due to foreground contamination. Data in this figure has
been smoothed to better show the line shapes, but all computations have been
carried out on the unsmoothed data. Note the
high-velocity component in both \ion{Si}{4} lines, extending out to
$\sim 850$~km~s$^{-1}$.}\label{fig:SingleLines}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3.500in]{./LyACoverfracs.pdf}
\caption{\textbf{Upper panel}: Ly$\alpha$ profile of Haro 11 C, in
approximate units of the surrounding continuum level. The full line
indicates the measured values smoothed by a 5 px. flat kernel; surrounding shading
encloses the $\pm 1 \sigma$ error band (mostly comparable in width to
the line width). \textbf{Middle panel}: Black steps show the averaged
LIS line profile, smoothed by a 5px kernel. Surrounding gray shading
denotes the $\pm 1 \sigma$ confidence band. \textbf{Lower panel}: Same
as middle panel, but for the \ion{Si}{4} transitions.
}\label{fig:HisLisLya}
\end{figure}

The picture is what we would expect from a LyC leaker: The neutral absorption
features are weak, and the ionized features are strong, revealing a   highly
ionized medium in front of the central cluster . This is also in good agreement
with the findings of \citet{Pardy2016arXiv}, who conclude that the galaxy has
about twice as much ionized as neutral gas. In addition to this, it is
interesting to note the close similarity in shape between the \ion{Si}{2} and
\ion{Si}{4} line profiles, indicating that they likely represent two different
phases in the same higher density regions. These regions will likely be ionized
on the side facing the central cluster, being photoionized by its hot, massive
stars. This also implies that the nebular emission outflow found in
Rivera-Thorsen et al, submitted, at least partially traces the same gas as the
\ion{Si}{4} absorption here and, by extension, also the neutral medium. Looking
at fig.~\ref{fig:SingleLines}, we see that maximum absorption and thus the
largest amount of gas in a single velocity bin for both phases is found at
around $v \approx -50$ km s⁻¹. Interestingly, a component was also found in
nebular emission at this velocity in the observations of Rivera-Thorsen et al.,
submitted. The outflow velocity $v_{\text{int}}$ is at 149 ± 7 km s⁻¹, fully
consistent with the findings of \citet{Heckman2015} and \citet{Alexandroff2015}.
Interestingly, \citet{Sandberg2013} found from the neutral Sodium resonance
absorption doublet $\lambda \lambda 5889.95,5895.92$ (NaD) an overall, weak
\emph{redshift} of $v = 32$ km s⁻¹. From e.g.~fig.~\ref{fig:HisLisLya}, it is
evident that while neutral gas \emph{is} present at these velocities, the
velocity is at odds with the integrated velocity of $v = -149$ km s⁻¹ found in
this work. We note that \ion{Na}{1} has a very low ionization potential of only
$\sim 5$ eV, meaning that these atoms may well only be present in the densest
and/or dustiest regions, in which Sodium is shielded from ionization.  This is
in agreement with the finding of \citet{Sandberg2013} that NaD, despite being a
strong transition, shows absorption of only $\sim 95\%$ of continuum level and
is mostly found in small, optically thick clouds.  This suggests that the NaD is
tracing only the densest and/or dustiest regions which are more slowly
accelerated by star formation feedback than the surrounding, more dilute medium.
In fig.~\ref{fig:apert}, dusty regions in the aperture are apparent E and W of
knot C. Seen in the better resolution of fig. 1 in \citet{Adamo2010}, these
regions seem like they might be connected by a narrow dust lane partially
covering the background source. We speculate that the NaD absorption of
\citet{Sandberg2013} might be associated with this.

Also the absorption feature in the Lyman $\alpha$ profile in the upper
panel seems to morphologically follow the shape of the metal lines,
indicating that radiative transfer effects are modest, indicative of a
fairly low column density of \ion{H}{1} around line center. We find a
Lyman $\alpha$ peak velocity of
$v_{\rm peak}^{\mathrm{Ly}\alpha} = 158 \pm 1$ km s⁻¹ relative to the 
systemic velocity found by \citet{Sandberg2013}, derived from nebular emission
lines in the region around knot C. This velocity, as it is also
discussed in \citet{Verhamme2015}, is just consistent with their
theoretical predictions for a density-bounded, low-column density
system, albeit on the upper limit of their allowed range.

@@

However, the strong riddled ionization bounded scenario -- i.e.
consisting of only a fully neutral and a fully ionized phase -- is easily ruled
out since the Ly$\alpha$ profile does not have any appreciable emission
component at zero velocity. We therefore expect a residual neutral fraction to
remain in the ionized phase; a fraction which has a column density high enough
to block Ly$\alpha$ efficiently at line center, but low enough to be at least
part transparent to LyC radiation and undetected in \ion{Si}{2}.

@@

The existence of a diffuse neutral component being present in the ionized medium
seems consistent with what is found in Lyman Break Analogs and Green Pea
galaxies \citep{Heckman2011, Henry2015}.  These galaxies have sometimes very low
LIS absorption depths, which is usually indicative of low covering fractions and
high porosity of the neutral medium - and yet they find covering fractions of
near unity of \ion{H}{1} from Lyman $\beta$ absorption, indicating column
densities of $N \gtrsim 10^{16}$ cm$^{-2}$. Like in the case of Haro 11, this
indicates that a non-negligible neutral component must be present in the ionized
phase, although it may very well be of such low densities and metallicity that
metal absorption from this medium is undetectable.

@@

In this work, we have re-analyzed an archival FUV HST-COS spectrum 
\textbf{of Haro 11} to
investigate the kinematics and geometry of both the hot, ionized and cold,
neutral ISM along the line of sight to a strongly star forming knot suspected of
being the source of the observed LyC leakege from the galaxy. We have used the
Apparent Optical Depth method to compute column densities and velocity-binned
covering fractions of the gas and compared these results to the extreme cases of
either an optically thin, density bounded neutral medium, or a riddled,
optically thick, ionization bounded neutral medium. Assuming that the LyC
emission previously observed from this galaxy indeed does originate from knot C,
we find that the observations are not compatible with the latter case, since the
characteristic, bright Lyman-$\alpha$ emission spike at line center is absent in
this spectrum. Furthermore, the observations are not consistent with an
optically thin, density bounded neutral medium.

@@

It is interesting to note that Haro 11 is relatively on neutral 
gas poor, and has a very strong ongoing starburst episode. If we were to see a 
case in which neutral gas depletion and/or ionization provided a 
density-bounded scenario, this would be a likely place to find it. 
Still, it seems that dynamical effects of feedback and galaxy-scale 
interaction plays the greater part in peeling away the neutral gas from the 
central starbursts and opening passages for Ly$\alpha$ and LyC to 
escape. Given that star formation and merger activity is significantly stronger 
at $z\gtrsim 1$ than in the local Universe, and that Dark Matter potential wells
were shallower, these mechanisms could be even more important in that era. 
one could speculate that this could provide a piece of the puzzle of which
sources have driven the epoch of reionization.
